---
title: "Exploratory Analysis"
output: html_document
---

## Access to Parks and Green Spaces in Seattle


### Authors

Kavya Rao, Maeve Zhu, Mina Trice

### Date

Spring 2024

### Abstract

For our project, we will analyze a dataset that contains data on every public park in the city of Seattle, specifically on what types of natural features and manmade amenities those parks contain. Our analysis will reveal details about the equal accessibility of the parks for citizens based on location and socioeconomic factors, what aspects of natural and recreational life are not facilitated in these parks, and even how wildlife-friendly the city is. This could be used to push the city government to enact change that makes its parks more accessible and useful for those who are not having their needs met.


### Keywords

Environment, parks & green spaces, accessibility

### Introduction  

*How common is access to parks and green spaces in Seattle? In particular, do Seattle parks tend to be accessible for people with disabilities?*

*What natural features and manmade amenities are most and least available in Seattle parks? What is the most accessible facility in Seattle parks?*

*Does the accessibility of parks and their important features vary by location within Seattle? Are the green areas concentrated in certain areas of the city or widely distributed?*

These questions are important because access to parks and green spaces for people in urban environments is proven to have a measurable impact on their mental and physical health. However, most cities were not planned and built with this in mind, so access to these green spaces can be skewed based on location. This, in turn, is determined by socioeconomic factors and historical practices such as redlining. Furthermore, the types of amenities the parks offer to users, as well as how well-maintained they are, can encourage or discourage residents from making full use of the parks. Residents require different types of parks, from child-friendly to dog-friendly to neither one, and so may not be able to utilize green spaces that do not fit their needs.


### Related Work  

In regards to parks and green spaces in urban areas, recent studies have further proved their importance, including locally in Seattle. A case study led by the University of Washington investigated Seattle’s biggest park, Discovery Park. When locals were surveyed on their most meaningful experience at the park, some of the main keywords included “watching birds”, “listening to waves at the beach”, “spotting wildlife”. This indicates that the wildness of Discovery Park, and urban parks in general, are important for one’s physical and mental wellbeing. (Lev et al., 2020)

 Lev, E., Kahn, P. H., Chen, H., & Esperum, G. (2020). Relatively wild urban parks can promote human resilience and flourishing: A case study of discovery park, Seattle, Washington. Frontiers in Sustainable Cities, 2. https://doi.org/10.3389/frsc.2020.00002 


### The Dataset

Where did you find the data? Please include a link to the data source

The data is collected by the Seattle Parks and Recreation Department. It was generated to provide detailed information about the features available at various parks within the city, aiding in park management and public information dissemination. [Access the dataset here](https://data.seattle.gov/Parks-and-Recreation/Seattle-Parks-and-Recreation-Parks-Features/2cer-njie/about_data) Data is provided by City of Seattle and the dataset owner is Bhang-Barnett, Hazel. There is no sufficient information on the website that mentions how the data was collected, but one of the columns contains qualitative data, such as features. We assume that the data of features and working hours are collected from on-site investigation. The city of Seattle also have access to the dataset of each park’s xpos and ypos. The data was collected because the Seattle Parks and Recreation Department wants to make data transparent and more accessible.This allows both the city officials and the public to better understand and access the resources. There are 1000 rows and 7 columns.

There are questions of power in the data. The data includes the feature, but doesn’t include the condition of the equipment. The parks belong to the government, thus the maintenance condition reflects the department’s effort put in maintaining a high-quality park. We think the government’s power may prevent them from showing the real status of the equipment.

Meanwhile, there are three possible limitations and problems with the data. In the first column there are several repetitive contents because the data is describing each park's features at a time. If a park has several features, the park’s name will appear several times in the first column. This means there are less than 1000 parks, most likely around 200-300 for the data analysis. The data includes the coordinates of the parks (latitude and longitude), but it does not include addresses or zip codes, which makes it more difficult to make a graph. This means that to analyze the data by location, we will have to make our own location groups through methods such as clustering certain latitudes and longitudes together. We will have to do more research on what location groups are relevant and how to imitate groups such as neighborhoods based only on the provided coordinates. One limitation includes the lack of data in regards to the quality of park features in Seattle. For example, the dataset includes “garden” as one of the features, but lacks the information regarding the quality of the garden. How big is the garden? Is it well maintained? Is it a garden for produce/community use, or simply one for show? The same lack of clarity applies to many other features such as tennis courts (How many courts are there?), hiking trails (How long is the trail?), views (What defines a “view”?) etc.

### Implications

Countless studies show that time in nature has tangible positive effects on mood, and that having access to parks also encourages more exercise and community building. All the people of Seattle deserve to have access to parks that facilitate this. Analyzing this dataset could show that certain amenities are clustered in some locations and not readily available in others. It could also show us that park location varies by social factors such as the income of certain areas. Furthermore, there are also implications about the biodiversity within the city, specifically whether it is wildlife-friendly. If too few parks are pesticide-free or contain hiking trails or waterfront, the park cannot be a safe and diverse wildlife corridor. This raises the question of whether enough parks are accessible for all citizens of Seattle as well as for wildlife.

In general, our findings could provide a direction for the city government as it prioritizes what aspects of its lands to work on. For example, they could possibly prioritize adding more greenery to parks without green features, or finding land for parks for people that do not have any access to them. Another example is that many of the provided features in the dataset relate to whether parks are child-friendly. This is important because child-friendliness encourages more people to make use of the park, and it is particularly important for children to have access to nature from a young age. Designers and policymakers could analyze this and choose to allocate funds to these features within Seattle’s parks.

### Limitations & Challenges

There are a few challenges and limitations in terms of our project. One limitation of this project includes the fact that this only focuses on parks in one city. We chose Seattle due to proximity and relevance to our group, however we cannot measure or generalize our findings to other cities in Washington, or across the world. As a result, our findings are limited to only Seattle. Additionally, another challenge that this project runs into is how these park features may look in the future, especially with climate change. Our project cannot predict and account for how things will change over the years, such as certain features shutting down, climate change affecting trails or waterfront areas, or maintenance quality of a park declining. Another limitation is that we are unable to see the user’s experience from this project idea. We can see some parks with fancy features, however, when you are traveling there you may find a lot of homeless people living near the park. Sometimes, there are people playing extremely loud music and doing drugs near the viewpoint. These are the unseen problems when visitors actually travel there. To build a broader and human-centric project, we should collect user experience data including qualitative data such as descriptive words based on their impression of the park.

### Summary Information

Firstly, we calculated the amount of occurrences of the features that appeared the most and least in the dataset. This showcased the discrepancy between them as well as allowing us to access the specific feature names for other purposes. We then calculated how many parks were open 24 hours, as having parks that stay open longer is important for accessibility. We also calculated the proportion of total features that were sports facilities, as having these facilities is crucial to encourage younger people to make use of the parks and we wanted to check whether these facilities were abundant. Finally, we calculated the proportion of play areas that were marked ADA compliant, because an important accessibility issue is that children who are disabled often have their needs that are specific to children overlooked.

### Table

As we discussed the importance of wild parks in urban spaces, this table below shows parks with said features. To do this, we sorted parks with the words "woods" and "hiking trails" in the features column. This showcases the almost 60 parks in Seattle with a less urban, and more wild environment.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(readr)
library(kableExtra)
final_table_data <- read_csv("final_table_data.csv")
final_table_data %>%
  kbl() %>%
  kable_styling()

```


### Chart 1

This scatterplot plots the distance of a park from the Seattle city center against how many different features the park has. It aims to provide a basic visualization of whether more varied and useful parks are more accessible to people in more central and urban or more peripheral and rural parts of Seattle.

```

parks_data = read.csv("Seattle_Parks_and_Recreation_Parks_Features_20240515.csv")

# Function to calculate distance of park from Seattle city center
calculate_distance <- function(lat1, lon1, lat2, lon2) {
  earth_radius <- 6371
  deg_lat <- deg2rad(lat2 - lat1)
  deg_lon <- deg2rad(lon2 - lon1)
  # Haversine formula
  a <- sin(deg_lat/2) * sin(deg_lat/2) + cos(deg2rad(lat1)) * cos(deg2rad(lat2)) * sin(deg_lon/2) * sin(deg_lon/2)
  c <- 2 * atan2(sqrt(a), sqrt(1-a))
  distance_km <- earth_radius * c
  return(distance_km)
}

# Function to convert degrees to radians
deg2rad <- function(deg) {
  return(deg * pi / 180)
}

# Seattle city center coordinates
city_center_lat <- 47.608013
city_center_lon <- -122.335167

# Calculate distance from Seattle city center for each park and make new column
parks_data$distance_from_center <- calculate_distance(city_center_lat, city_center_lon, parks$yPos, parks$xPos)

# Count how many features each park has and make new column
parks_data$feature_count <- parks$Name %>% str_count()

# Make scatterplot
distance_scatterplot <- ggplot(parks, aes(x = feature_count, y = distance_from_center)) +
  geom_point() +
  labs(x = "Number of Features", y = "Distance from City Center (km)") +
  ggtitle("Scatterplot of Park Features and Distance from City Center")

# Print scatterplot
print(distance_scatterplot)

```

### Chart 2

Include a chart

Describe why you chose this chart and what information it reveals

### Chart 3

This interactive map showcases the number of parks in Seattle with ADA compliant amenities. This includes play areas, paths and restrooms, with some parks including all three. As we discussed earlier, the importance of accessible features in parks are crucial for everyone's enjoyment and equal accessibility to amenities. It is also important to be mindful of how we can go beyond these 128 parks w/ ADA compliant features.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(leaflet)
library(shiny)
library(dplyr)

final_ada_data <- read.csv("final_ada_data.csv")


leaflet(data = final_ada_data) %>%
  addTiles(options = providerTileOptions(noWrap = TRUE)) %>%
  addMarkers(lng = ~xPos, lat = ~yPos, popup = ~paste("<b>", Name, "</b><br>", Feature_Desc))
```

